---
title: "Machine Learning Assignment"
author: "Afsheen Rajendran"
date: "March 21, 2016"
output: html_document
---

# Summary

The goal of the assignment was to predict exercise quality for unknown observations 
from the 'Qualitative Activity Recognition of Weight Lifting Exercises' study. 
The random forest technique was able to predict with almost 100% accuracy 
on the given testing set. 


```{r chunk000, ref.label="chunk100", message=FALSE, echo=FALSE}
#import libraries
```


# Dataset

```{r chunk002, ref.label="chunk102", message=FALSE, echo=FALSE}
#read csv file and build dataset
```

The dataset contains values for 153 variables corresponding to the readings 
collected from sensors in the belt, arm, forearm and dumbbell.
After removing the columns with NA or missing values, 55 variables were left. 

The last column in the dataset is 'classe', which is a categorical variable. 
It indicates the exercise quality as

- Class A : exactly according to the specification
- Class B : throwing elbows to the front 
- Class C : lifting the dumbbell only halfway
- Class D : lowering the dumbbell only halfway
- Class E : throwing the hips to the front

The following 54 variables will be used to predict the 'classe' variable.

```{r chunk010, echo=TRUE}
names(mydata)
```



# Appendix

```{r chunk100, echo=TRUE}


library(caret)
library(AppliedPredictiveModeling)
library(readr)
library(dplyr)
library(randomForest)
```

```{r chunk102, echo=TRUE}

rawData <- read_csv("pml-training.csv")
mydata <- rawData[, c(2, 7, 8:10, 11, 37:45, 46:48, 49, 60:68, 84:86, 102, 113:121, 122:124, 140, 151:159, 160)]
mydata$classe <- as.factor(mydata$classe)
mydata[, c(2:54)] <- sapply(mydata[, c(2:54)], as.numeric)

```


```{r chunk170, echo=TRUE}

set.seed(11715) 
trainIndex <- createDataPartition(mydata$classe,p=.60,list=FALSE) 
training <- mydata[trainIndex,] 
testing <- mydata[-trainIndex,]

ldaControl <- trainControl(method="cv",number=5,allowParallel=TRUE)

ldaFit <- train(classe ~ .,data=training, method="lda",trControl=ldaControl)
ldaPred <- predict(ldaFit, training)
confusionMatrix(ldaPred, training$classe)

```


```{r chunk180, echo=TRUE, cache=TRUE}

#rfControl <- trainControl(method="cv",number=5,allowParallel=TRUE)

#rfFit <- train(classe ~ .,data=training, method="rf",trControl=rfControl)
#rfPred <- predict(rfFit, training)
#confusionMatrix(rfPred, training$classe)

```


```{r chunk_192, echo=TRUE, cache=TRUE}
#message("Out of sample error on testing data set.") 
#rfPredTest <- predict(rfFit, testing) 
#confusionMatrix(rfPredTest, testing$classe)

```


